The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-02-04 Paper/jaccardbaetica.tex
add changes germania brit

@@

We analysed a dataset of 2219 stamps from different centres in Britannia. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
The stamp database was also compiled by CEIPAC database. The database contained anomalies that were eliminated using the same criteria cited above. We finally selected the centres with more or equal than 5 stamps. As a result, we studied a total of 1765 stamps comprised of 968 different stamps from 46 centres.
The centres can be seen in Fig.\ref{britannia}.

@@

@@

%\xavi{este titulín es mejorable :-D}

@@

@@

%\xavi{no pillo demasiado lo que comentas de "groupings of similar stamps were not found in the cluster".}
%\xavi{Si usas esos colores aquí y en la figura anterior entonces deberías usarlos también en el mapa.}

%%\subsection{Jaccard distance}

%Result of Jaccard distance can be The coefficients range from x to x.The dendrogram shown in Fig. Bla bla was obtained with the Jaccard distance measure. The dendrogram suggests

@@

\bibliography{bibliotex}

>>>>

The consumption of olive oil in Britannia was residual until the start of the Roman conquest \citep{funari_corpus_1996,
carreras_abastecimiento_2003}.

It is well known that this product was not frequently consumed by autonomous population. The absence of olive oil importation before the conquest can be reflected by the shortage of this product until the arrival of high percentage of Dressel 20 amphorae to Britannia \citep[ 1]{carreras_britannia_1998}. Additionally, the land of Britannia showed inadequate for the olive oil production due to low environmental conditions. This issue was necessary for creating an important export apparatus from Baetica to supply Roman legions.
In this moment, we detect an increase of the olive oil exportation concurring with the displacement of Roman legions during the military campaigns \citep[ 161]{monfort_britanniaen_1998}. This fact will be spatial incidence in sites close at Hadrian Wall.  
Olive oil production in Baetica would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns. 
%debate intercambios puntuales o redes comerciales de abastecimiento

The presence of Dressel 20 stamps in military camps in Britannia have been widely studied in Roman archaeology \citep{carreras_britannia_1998}. This fact also indicates a possible governmental organisation allocate to reforce and supply of olive oil in military camps, such as Germania (Remesal Rodrıguez, 1986). However, it is unknown the type of economical system when managing or control the redistribution of olive oil in different sites in Britannia but it seems that it could have been in the main cities \citep[45]{funari_economic_2005}.

Finally, The increase in olive oil exports would experience a progressive slowdown
from the third century A.D., coinciding with the change in market strategy in the
Empire. In that moment, a progressive dicrease of Dressel 20 is documented being gradually replace for Dressel 23 amphorae.

@@

We analysed a dataset of 2219 stamps from different centres in Britannia. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
The stamp database was also compiled by CEIPAC database. The database contained anomalies that were eliminated using the same criteria cited above. We finally selected the centres with more or equal than 5 stamps. As a result, we studied a total of 1765 stamps comprised of 968 different stamps from 46 centres.
The centres located in Britannia can be seen in Fig.\ref{britannia}.

@@

\subsubsection{Germania}

The Roman conquest in Germania dates from the end of the 1st century BC, during Caesar campaign and subsequently continued by Augusto. The advance of the Roman legions toward Gallia opened a way though the Atlantic Ocean to Germania \citep{remesal_annona_1986,
remesal_baetica_2002}.

It is thought that this route would have been initiated previously through 
terrestrial transport, although recent studies question this position giving an important role to the Atlantic route for both the conquests of Germania and those of Britannia \citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

As the case of Britain, the consumption of olive oil in Germania was not detected before the conquest.

The need to supply roman legions created a logistic network to supply to the army. In this moment, we detect the arrival of olive oil amphorae to Germania evidenced by the presence of Dressel 20 amphora in different sites close to limes.

The roman studies about the presence of Dressel 20 amphorae has not been the same repercusion than the rest of the province. This can be explained due to the lack of data \citep{horacio2010llegada}.

As a consequence, we do not the role that commerce agents in the participation of the distribution and exportation of olive oil in Germania or, on the contrary, the Roman army was the responsible to supply the product in the own province \citep[156]{remesal_germn_2010}.

The presence of Roman army encouraged the exchange in the province showed by the arrival of this product both civil settlement and military sites with a mayor concentration at german limes. It seems that some baetican centres would be assigned to the support of olive oil. However, this hypothesis can not be proven by the lack of archaeological sources \citep[125]{remesal_concierto}.

We analysed a dataset of 2052 stamps placed in Germania. All the data was also compiled by CEIPAC database. We finally selected a total of 1621 stamps consist of 850 different stamps from 46 sites (see Fig. \ref{germania}). As previously before, the centres were selected with more or equal than 5 stamps.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania5.pdf}
\caption{Distribution of the sites in Germania. Most sites were located in limes during the Roman Empire}
\label{germania}
\end{figure}

@@

@@

\subsection{Production centres: Baetica Province}

@@

\subsection{Reception centres: Britannia and Germania}

The results show a correlation similar to the obtained in Baetica province. However, the similarity is less pronounced as we observed in the dendrogram in Fig. \ref{britmap}

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorthm of different sites in Britannia. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{britmap}
\end{figure}

A minor correlation can be explained by the fact that Britannia spatial area is much wide than \textit{Baetica} where the centres were more concentrated.

Most military sites that showed greater similarity in the results were geographically close. We identified two factors in the results: 1) nearby sites tend to share similar stamps and 2) most sites have different stamps. 
By contrast, most sites did not show a strong similarity in stamp correlation . Neither we found a grouping of similar stamps in a specific place. In general, we did not observe any production pattern that indicates a clear organisation between production centres and reception centres. In other words, our results did not indicate the presence of a specific production center from Baetica in Britannia province.  Rather BLA BLA

\subsection{Germania}

The similarity of stamps show a results less significative than Britannia province as it can be seen in the dendrogram Fig. \ref{germap}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorthm of different sites in Germania. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{figure}

The sites with more similarity in stamps were closely located in limes. In this case, most sites have different stamps  which means that we did not detect any pattern. On the other hand, there seems to be a greater concentration of the seals in areas eminently militarized and close to the german limes. However, It is still no possible to determine with the data that the existence of a defined pattern that reflects in more detail the route of this production in the area of the German limes.

@@

%\bibliography{bibliotex}
\bibliography{bibtesis}
